---
title: "MortaLIVE analysis workflow"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
params:
  run_viability: false
  run_bliss: false
  run_outgrowth: false
  run_ic50: false
  run_validation: false
  run_comparison: false
  run_sholl_assembly: false
---

# Overview

This document combines the MortaLIVE paper's analysis workflows into one executable
notebook. The sections cover viability normalization, Bliss independence, exposure-related
outgrowth, IC50 estimation, assay compatibility validation, and method comparisons.
An optional Fiji CSV assembly utility supports profile preparation.

| Manuscript panel | Analysis section |
| --- | --- |
| Fig. 1d–e | Compatibility validation: normalized β-III-tubulin and GFAP AUC |
| Fig. S6 | Bliss independence: single agents versus mixture |
| Fig. 3c | Method comparison: mean intersection traces |
| Fig. 3d | Method comparison: raw AUC versus MortaLIVE |
| Fig. S7; Fig. 4b–c | Exposure-related Sholl AUC analysis; configure GFAP, NO, or ML inputs |
| Fig. 4d | Outgrowth concentration–response fitting and IC50 intervals |
| Upstream viability processing | Within-experiment VC normalization and workbook combination |

Configure input files in **Configuration**, then enable the corresponding parameters
in the YAML header or in `rmarkdown::render()`. All analyses are disabled initially,
so the document can render its methods and code without experimental input files.
The sections use independent inputs; enabling one does not automatically enable the
others. Outgrowth IC50 inputs may be generated in the same run by the outgrowth section.

The workflow preserves the analytical choices in the source scripts, with input
validation and implementation corrections documented in `README.md`. It does not
perform image segmentation, plate assignment, longitudinal ANOVA, or cross-assay
paired comparisons. Raw images and experimental data are not embedded in this file.

```{r setup}
# Also permits sequential interactive execution outside rmarkdown::render().
if (!exists("params")) {
  params <- list(run_viability = FALSE, run_bliss = FALSE,
                 run_outgrowth = FALSE, run_ic50 = FALSE, run_validation = FALSE,
                 run_comparison = FALSE, run_sholl_assembly = FALSE)
}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                     error = FALSE, fig.width = 12, fig.height = 5)
options(stringsAsFactors = FALSE)
required <- c("knitr", "rmarkdown")
if (params$run_viability) required <- c(required, "openxlsx")
if (params$run_bliss) required <- c(required, "readxl", "dplyr", "tidyr",
                                  "tibble", "janitor", "writexl")
if (params$run_outgrowth || params$run_validation || params$run_comparison) required <- c(required, "readxl", "readr", "dplyr",
    "tidyr", "tibble", "purrr", "stringr", "MESS", "ggplot2", "ggpubr", "writexl")
if (params$run_ic50) required <- c(required, "readr", "dplyr", "tidyr",
                                 "tibble", "purrr", "minpack.lm", "writexl", "MASS")
if (params$run_sholl_assembly) required <- c(required, "dplyr", "stringr", "writexl")
required <- unique(required)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Install required packages before running: ",
                          paste(missing, collapse = ", "))
# Only packages needed by enabled sections are attached. Installation is separate.
invisible(lapply(setdiff(required, c("MASS", "knitr", "rmarkdown")),
                 library, character.only = TRUE))
```

# Configuration

Paths are relative to the directory containing this R Markdown file. Each run ID
identifies a source experiment, not a biological replicate or a statistical block.
Do not list the same experiment twice under different IDs.

```{r configuration}
output_root <- "analysis_output"
overwrite_outputs <- FALSE
QC_TOLERANCE <- 1e-8

# 1. Viability: add any number of independent source workbooks.
viability_files <- c(run01 = "data/viability_run01.xlsx",
                     run02 = "data/viability_run02.xlsx",
                     run03 = "data/viability_run03.xlsx")
analysis_sheet_pattern <- "^(Cd|Cr|Cd\\+Cr|Pb) Norm (LD|LIVE|DEAD)$"
treatment_order <- c("VC", "0.32 µM", "3.2 µM", "10 µM", "50 µM", "Death Control")
# NULL retains every timepoint. Set 72 to exclude later observations explicitly.
viability_max_time_h <- NULL

# 2. Bliss: one row per matched component concentration, with replicate columns.
bliss_input_file <- "data/mixture_72h.xlsx"
bliss_input_sheet <- "RME quantification"
bliss_bootstrap_iterations <- 5000L
bliss_seed <- 831L

# 3. Outgrowth: one row per assay and independently normalized experiment.
# All profiles in a selected sheet must belong to the same source experiment.
outgrowth_inputs <- data.frame(
  Assay = rep(c("ML", "NO"), each = 3),
  Run_ID = rep(c("run01", "run02", "run03"), times = 2),
  File = paste0("data/", rep(c("ML", "NO"), each = 3), "_",
                rep(c("run01", "run02", "run03"), times = 2), ".xlsx"),
  Sheet = "Sheet1"
)
outgrowth_radius_unit <- "um"
outgrowth_radius_step <- NULL  # No subsampling for already-formatted profiles.
outgrowth_exact <- FALSE       # Use NULL for the earlier acute-metal script's automatic exact-test choice.
outgrowth_doses_uM <- c(0, 0.32, 3.2, 10)
outgrowth_metals <- c("Cd", "Cr", "Mix", "Pb")
single_run_min_n <- 1L
pooled_min_n <- 2L
wilcoxon_adjustment <- "BH"
single_run_display_cutoff <- 0.075
pooled_display_cutoff <- 0.05
plot_error <- "SEM"  # Source scripts use SEM; SD is an explicit alternative.
plot_seed <- 831L

# 4. Compatibility validation (Fig. 1d–e): separate marker datasets.
validation_inputs <- data.frame(
  Marker = c("btubIII", "GFAP"), Source_ID = c("btubIII_validation", "GFAP_validation"),
  File = c("data/validation_btubIII.xlsx", "data/validation_GFAP.xlsx"), Sheet = "Sheet1"
)
validation_groups <- c("Ctrl", "0.0005ML", "0.001ML")
validation_radius_step <- 10  # Retain multiples of 10 input pixels, as in scripts 06–07.
validation_radius_unit <- "px"
validation_radius_scale <- 1 # For conversion, supply calibration and change the output unit label.

# 5. Method comparison (Fig. 3c–d): add a measured combined-marker row if available.
comparison_inputs <- data.frame(
  Method = c("IF_bTubIII", "IF_GFAP", "ML"),
  Source_ID = c("IF_bTubIII", "IF_GFAP", "ML"),
  File = c("data/comparison_IF.xlsx", "data/comparison_IF_GFAP.xlsx", "data/comparison_ML.xlsx"),
  Sheet = "Sheet1"
)
# Optional row: Method = "IF_bTubIII_GFAP", with its own measured profile workbook.
comparison_reference <- "ML"
comparison_radius_step <- 10
comparison_radius_unit <- "px"
comparison_radius_scale <- 1 # Filtering occurs in input units before this conversion.
comparison_ribbon <- "SEM" # Source default; CI95_half_width is an explicit alternative.

# 6. Optional CSV preprocessing; independent of the figure analyses.
sholl_parent_dir <- "data/sholl_profiles"

# 7. IC50: pooled outgrowth CSVs, either existing or generated in section 3.
ic50_files <- c(
  ML = file.path(output_root, "outgrowth", "ML_combined_normalized_AUC_per_replicate.csv"),
  NO = file.path(output_root, "outgrowth", "NO_combined_normalized_AUC_per_replicate.csv")
)
ic50_metals <- c("Cd", "Cr", "Mix", "Pb")
max_iterations <- 1000L
max_IC50 <- 1000
max_logIC50 <- log10(max_IC50)
```

## Shared validation

Blank measurements remain missing. Non-numeric entries, duplicated identifiers,
invalid denominators, and incompatible input layouts stop execution with a message.

```{r validation-functions}
check_file <- function(path, sheet = NULL) {
  if (!file.exists(path)) stop("Input file not found: ", path)
  if (!is.null(sheet) && !sheet %in% readxl::excel_sheets(path))
    stop("Worksheet '", sheet, "' not found in ", path)
  invisible(path)
}
strict_numeric <- function(x, context) {
  text <- trimws(as.character(x))
  blank <- is.na(x) | text == ""
  value <- suppressWarnings(as.numeric(text))
  if (any(!blank & !is.finite(value)))
    stop("Non-numeric or non-finite value in ", context, ": ",
         paste(unique(text[!blank & !is.finite(value)]), collapse = ", "))
  value[blank] <- NA_real_
  value
}
output_path <- function(section, filename) {
  directory <- file.path(output_root, section)
  dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  path <- file.path(directory, filename)
  if (file.exists(path) && !overwrite_outputs)
    stop("Output exists: ", path, ". Choose a new output_root or enable overwrite_outputs.")
  path
}
check_ids <- function(ids, context) {
  if (!length(ids) || anyNA(ids) || any(!grepl("^[A-Za-z0-9_-]+$", ids)) ||
      anyDuplicated(ids)) stop(context, " must contain unique, nonempty IDs using letters, digits, _ or -.")
}
```

# Viability normalization and combination

## Input and calculation

Each workbook contains two-header-row sheets named, for example, `Cd Norm LD`,
`Cr Norm LIVE`, or `Cd+Cr Norm DEAD`. Row 1 identifies treatment blocks, row 2
contains well IDs, and column A contains time in hours. The treatment label appears
only at the start of its block. Blank replicate slots must retain their columns.

For each source experiment, endpoint, and timepoint:

\[
Y_{\%VC}=100\,Y/\overline{Y}_{VC}.
\]

These inputs are already image-derived measurements; the code does not calculate
organoid area or perform live/dead segmentation. Individual VC values vary around
100%; only their mean is fixed. Combination preserves replicate-level values and
matches rows by time. It does not average experiments or repeat normalization.

```{r viability-normalization-functions}
read_mortalive_sheet <- function(file, sheet) {
  raw <- openxlsx::read.xlsx(
    xlsxFile = file,
    sheet = sheet,
    colNames = FALSE,
    skipEmptyRows = FALSE,
    skipEmptyCols = FALSE,
    check.names = FALSE
  )
  if (nrow(raw) < 3) {
    stop(sprintf("Sheet '%s' does not contain two header rows plus data.", sheet))
  }
  raw
}
# Detect the contiguous VC block from header row 1.
find_vc_columns <- function(header_row, sheet) {
  header_row <- trimws(as.character(header_row))
  vc_start <- which(!is.na(header_row) & header_row == "VC")
  if (length(vc_start) != 1) {
    stop(sprintf(
      "Expected exactly one 'VC' header in sheet '%s'; found %d.",
      sheet,
      length(vc_start)
    ))
  }
  nonblank_headers <- which(
    !is.na(header_row) & nchar(trimws(header_row)) > 0
  )
  next_group_start <- nonblank_headers[nonblank_headers > vc_start][1]
  if (is.na(next_group_start)) {
    stop(sprintf(
      "Could not identify the treatment block following VC in sheet '%s'.",
      sheet
    ))
  }
  vc_start:(next_group_start - 1)
}
# Normalize one analysis sheet to its timepoint-matched VC mean.
normalize_sheet_to_vc <- function(file, sheet, tolerance = QC_TOLERANCE) {
  raw <- read_mortalive_sheet(file, sheet)
  # First row = treatment headers; second row = well IDs.
  header1 <- unlist(raw[1, , drop = TRUE], use.names = FALSE)
  # Rows 3 onward contain time and numeric measurements.
  dat <- raw[-c(1, 2), , drop = FALSE]
  # Convert the data block to numeric.
  # Blank spacer cells remain NA.
  dat[] <- lapply(dat, function(x) {
    strict_numeric(x, paste(basename(file), sheet))
  })
  # Keep only actual timepoint rows.
  missing_time <- is.na(dat[[1]])
  if (any(missing_time & rowSums(!is.na(dat[-1])) > 0))
    stop("Measurements without a numeric timepoint in ", sheet)
  dat <- dat[!missing_time, , drop = FALSE]
  if (anyDuplicated(dat[[1]])) stop("Duplicate timepoints in ", sheet)
  if (nrow(dat) == 0) {
    stop(sprintf("No numeric timepoints found in sheet '%s'.", sheet))
  }
  vc_cols <- find_vc_columns(header1, sheet)
  # Mean VC at each timepoint. Missing VC wells are ignored, NOT treated as zero.
  vc_matrix <- as.matrix(dat[, vc_cols, drop = FALSE])
  vc_n <- rowSums(!is.na(vc_matrix))
  vc_mean <- rowMeans(vc_matrix, na.rm = TRUE)
  # Guard against invalid normalization denominators.
  bad_vc <- !is.finite(vc_mean) | vc_mean <= 0 | vc_n == 0
  if (any(bad_vc)) {
    bad_times <- paste(dat[[1]][bad_vc], collapse = ", ")
    stop(sprintf(
      "Invalid VC mean in sheet '%s' at timepoint(s): %s",
      sheet,
      bad_times
    ))
  }
  normalized <- dat
  # Column 1 is TIME and must not be normalized.
  measurement_cols <- 2:ncol(normalized)
  normalized_matrix <- as.matrix(
    normalized[, measurement_cols, drop = FALSE]
  )
  normalized[, measurement_cols] <- sweep(
    normalized_matrix,
    MARGIN = 1,
    STATS = vc_mean,
    FUN = "/"
  ) * 100
  # QC: the mean of the available normalized VC wells must be 100 at every timepoint.
  normalized_vc_mean <- rowMeans(
    as.matrix(normalized[, vc_cols, drop = FALSE]),
    na.rm = TRUE
  )
  max_abs_error <- max(abs(normalized_vc_mean - 100), na.rm = TRUE)
  if (max_abs_error > tolerance) {
    stop(sprintf(
      "VC normalization QC failed for sheet '%s'. Max absolute error = %.12g",
      sheet,
      max_abs_error
    ))
  }
  list(
    data = normalized,
    vc_cols = vc_cols,
    vc_mean_raw = vc_mean,
    vc_n = vc_n,
    normalized_vc_mean = normalized_vc_mean,
    max_abs_error = max_abs_error
  )
}
# Normalize every MortaLIVE analysis sheet in one workbook.
# The original workbook is loaded and only the numeric data rows are replaced,
# so the existing worksheet structure, headers, and Well Map are retained.
normalize_workbook_to_vc <- function(input_file, output_file) {
  if (!file.exists(input_file)) {
    stop(sprintf("Input file not found: %s", input_file))
  }
  all_sheets <- openxlsx::getSheetNames(input_file)
  analysis_sheets <- grep(
    analysis_sheet_pattern,
    all_sheets,
    value = TRUE
  )
  if (length(analysis_sheets) == 0) {
    stop(sprintf("No MortaLIVE analysis sheets found in: %s", input_file))
  }
  wb <- openxlsx::loadWorkbook(input_file)
  qc_rows <- vector("list", length(analysis_sheets))
  for (i in seq_along(analysis_sheets)) {
    sheet <- analysis_sheets[i]
    result <- normalize_sheet_to_vc(input_file, sheet)
    # Write normalized numeric values back beginning at Excel row 3.
    # Rows 1-2 and workbook formatting are left in place.
    source_raw <- read_mortalive_sheet(input_file, sheet)
    openxlsx::deleteData(wb, sheet, cols = seq_len(ncol(source_raw)),
                         rows = seq.int(3, nrow(source_raw)), gridExpand = TRUE)
    openxlsx::writeData(
      wb = wb,
      sheet = sheet,
      x = result$data,
      startCol = 1,
      startRow = 3,
      colNames = FALSE,
      rowNames = FALSE,
      keepNA = FALSE
    )
    qc_rows[[i]] <- data.frame(
      sheet = sheet,
      timepoints = nrow(result$data),
      vc_n_min = min(result$vc_n),
      vc_n_max = max(result$vc_n),
      max_abs_error_from_100 = result$max_abs_error,
      stringsAsFactors = FALSE
    )
  }
  openxlsx::saveWorkbook(
    wb,
    file = output_file,
    overwrite = TRUE
  )
  do.call(rbind, qc_rows)
}
```

```{r viability-combination-readers}
# Clean header cells
clean_header <- function(x) {
  x <- as.character(x)
  x[is.na(x)] <- ""
  trimws(x)
}
# Identify the six treatment blocks
find_treatment_blocks <- function(
  raw,
  sheet
) {
  header1 <- clean_header(
    unlist(
      raw[1, , drop = TRUE],
      use.names = FALSE
    )
  )
  if (any(vapply(treatment_order, function(x) sum(header1 == x), integer(1)) != 1L))
    stop("Each configured treatment must occur exactly once in row 1 of ", sheet)
  starts <- match(
    treatment_order,
    header1
  )
  if (anyNA(starts)) {
    missing_treatments <- treatment_order[
      is.na(starts)
    ]
    stop(
      sprintf(
        "Sheet '%s' is missing treatment header(s): %s",
        sheet,
        paste(
          missing_treatments,
          collapse = ", "
        )
      )
    )
  }
  # Confirm treatments occur in the expected order
  if (any(diff(starts) <= 0)) {
    stop(
      sprintf(
        "Treatment blocks are not in the expected order in sheet '%s'.",
        sheet
      )
    )
  }
  # Each treatment extends until the column before
  # the next treatment header.
  ends <- c(
    starts[-1] - 1,
    ncol(raw)
  )
  blocks <- Map(
    seq.int,
    starts,
    ends
  )
  names(blocks) <- treatment_order
  widths <- lengths(blocks)
  # All treatment blocks in one source workbook
  # should have equal capacity.
  if (length(unique(widths)) != 1) {
    stop(
      sprintf(
        "Treatment blocks have unequal widths in sheet '%s': %s",
        sheet,
        paste(
          widths,
          collapse = ", "
        )
      )
    )
  }
  list(
    blocks = blocks,
    block_width = widths[1]
  )
}
# Read one normalized analysis sheet from one batch
read_normalized_batch_sheet <- function(
  file,
  sheet,
  batch_tag
) {
  raw <- read_mortalive_sheet(
    file = file,
    sheet = sheet
  )
  block_info <- find_treatment_blocks(
    raw = raw,
    sheet = sheet
  )
  # Row 2 contains well IDs
  header2 <- clean_header(
    unlist(
      raw[2, , drop = TRUE],
      use.names = FALSE
    )
  )
  # Rows 3 onward contain time and numerical data
  dat <- raw[
    -c(1, 2),
    ,
    drop = FALSE
  ]
  dat[] <- lapply(dat, strict_numeric, context = paste(basename(file), sheet))
  if (any(is.na(dat[[1]]) & rowSums(!is.na(dat[-1])) > 0))
    stop("Measurements without a timepoint in ", sheet)
  # Remove any rows without a numeric timepoint
  dat <- dat[
    !is.na(dat[[1]]),
    ,
    drop = FALSE
  ]
  if (!is.null(viability_max_time_h))
    dat <- dat[dat[[1]] <= viability_max_time_h, , drop = FALSE]
  times <- as.numeric(
    dat[[1]]
  )
  if (length(times) == 0) {
    stop(
      sprintf(
        "No numeric timepoints found in sheet '%s' of '%s'.",
        sheet,
        basename(file)
      )
    )
  }
  if (anyDuplicated(times)) {
    stop(
      sprintf(
        "Duplicate timepoints found in sheet '%s' of '%s'.",
        sheet,
        basename(file)
      )
    )
  }
  well_ids <- vector(
    "list",
    length(treatment_order)
  )
  values <- vector(
    "list",
    length(treatment_order)
  )
  names(well_ids) <- treatment_order
  names(values) <- treatment_order
  for (treatment in treatment_order) {
    cols <- block_info$blocks[[treatment]]
    wells <- header2[cols]
    populated <- colSums(!is.na(dat[, cols, drop = FALSE])) > 0
    if (any(populated & wells == "")) stop("Measurements have no well ID in ", sheet)
    if (anyDuplicated(wells[wells != ""])) stop("Duplicate well ID within a treatment in ", sheet)
    # Prefix actual well IDs with the batch number.
    #
    # Example:
    # A2 -> b1_A2
    #
    # Existing blank replicate slots remain blank.
    wells <- ifelse(
      wells == "",
      "",
      paste0(
        batch_tag,
        "_",
        wells
      )
    )
    well_ids[[treatment]] <- wells
    values[[treatment]] <- as.matrix(
      dat[
        ,
        cols,
        drop = FALSE
      ]
    )
  }
  list(
    time = times,
    well_ids = well_ids,
    values = values,
    block_width = block_info$block_width
  )
}
```

```{r viability-combination-function}
combine_one_analysis_sheet <- function(
  files,
  batch_tags,
  sheet
) {
  # Read the same sheet from every batch
  batch_data <- Map(
    function(file, tag) {
      read_normalized_batch_sheet(
        file = file,
        sheet = sheet,
        batch_tag = tag
      )
    },
    files,
    batch_tags
  )
  # the first experiment establishes the reference time order
  reference_times <- batch_data[[1]]$time
  # Match every batch to the first experiment by actual time value
  for (i in seq_along(batch_data)) {
    current_times <- batch_data[[i]]$time
    missing_times <- setdiff(
      reference_times,
      current_times
    )
    extra_times <- setdiff(
      current_times,
      reference_times
    )
    if (
      length(missing_times) > 0 ||
      length(extra_times) > 0
    ) {
      stop(
        sprintf(
          paste0(
            "Timepoint mismatch in sheet '%s' for %s. ",
            "Missing: [%s]; Extra: [%s]"
          ),
          sheet,
          batch_tags[i],
          paste(
            missing_times,
            collapse = ", "
          ),
          paste(
            extra_times,
            collapse = ", "
          )
        )
      )
    }
    time_match <- match(
      reference_times,
      current_times
    )
    if (anyNA(time_match)) {
      stop(
        sprintf(
          "Could not time-match %s in sheet '%s'.",
          batch_tags[i],
          sheet
        )
      )
    }
    # Reorder numerical data to the reference time order
    for (treatment in treatment_order) {
      batch_data[[i]]$values[[treatment]] <-
        batch_data[[i]]$values[[treatment]][
          time_match,
          ,
          drop = FALSE
        ]
    }
  }
  # Determine how many replicate slots each batch contributes
  source_widths <- vapply(
    batch_data,
    function(x) {
      x$block_width
    },
    numeric(1)
  )
  # Sum replicate capacities across all source experiments.
  combined_block_width <- sum(
    source_widths
  )
  # Begin combined table with TIME
  header1 <- "TIME (h)"
  header2 <- ""
  combined_values <- matrix(
    reference_times,
    ncol = 1
  )
  # Combine one treatment at a time
  for (treatment in treatment_order) {
    # Row 1:
    # treatment name followed by blank subheaders
    treatment_header <- c(
      treatment,
      rep(
        "",
        combined_block_width - 1
      )
    )
    # Row 2:
    # b1 wells followed by b2 wells, etc.
    treatment_wells <- unlist(
      lapply(
        batch_data,
        function(x) {
          x$well_ids[[treatment]]
        }
      ),
      use.names = FALSE
    )
    # Numerical values:
    # the first experiment columns followed by the next experiment columns
    treatment_values <- do.call(
      cbind,
      lapply(
        batch_data,
        function(x) {
          x$values[[treatment]]
        }
      )
    )
    # Width QC
    if (
      length(treatment_wells) !=
      combined_block_width
    ) {
      stop(
        sprintf(
          "Well-header width error for '%s' in sheet '%s'.",
          treatment,
          sheet
        )
      )
    }
    if (
      ncol(treatment_values) !=
      combined_block_width
    ) {
      stop(
        sprintf(
          "Data width error for '%s' in sheet '%s'.",
          treatment,
          sheet
        )
      )
    }
    header1 <- c(
      header1,
      treatment_header
    )
    header2 <- c(
      header2,
      treatment_wells
    )
    combined_values <- cbind(
      combined_values,
      treatment_values
    )
  }
  # Final expected number of columns
  expected_columns <-
    1 +
    length(treatment_order) *
    combined_block_width
  if (
    ncol(combined_values) !=
    expected_columns
  ) {
    stop(
      sprintf(
        paste0(
          "Combined column-count QC failed for sheet '%s'. ",
          "Expected %d columns, found %d."
        ),
        sheet,
        expected_columns,
        ncol(combined_values)
      )
    )
  }
  list(
    header1 = header1,
    header2 = header2,
    data = combined_values,
    block_width = combined_block_width,
    source_widths = source_widths,
    timepoints = reference_times,
    batch_data = batch_data
  )
}
```

```{r viability-copy-qc}
verify_combined_sheet <- function(
  result,
  sheet,
  batch_tags
) {
  combined_width <- result$block_width
  source_widths <- result$source_widths
  for (t in seq_along(treatment_order)) {
    treatment <- treatment_order[t]
    # First combined column belonging to this treatment.
    #
    # Column 1 is TIME, so the first treatment begins at 2.
    treatment_start <-
      2 +
      (t - 1) *
      combined_width
    batch_offset <- 0
    for (b in seq_along(result$batch_data)) {
      width_b <- source_widths[b]
      output_cols <-
        treatment_start +
        batch_offset +
        seq_len(width_b) -
        1
      source_values <-
        result$batch_data[[b]]$values[[treatment]]
      output_values <-
        result$data[
          ,
          output_cols,
          drop = FALSE
        ]
      same_values <- isTRUE(
        all.equal(
          source_values,
          output_values,
          check.attributes = FALSE,
          tolerance = 0
        )
      )
      if (!same_values) {
        stop(
          sprintf(
            "Copy QC failed for %s, '%s', sheet '%s'.",
            batch_tags[b],
            treatment,
            sheet
          )
        )
      }
      batch_offset <-
        batch_offset +
        width_b
    }
  }
  TRUE
}
```

```{r viability-workbook-writer}
combine_normalized_workbooks <- function(
  files,
  batch_tags,
  output_file
) {
  # Use first batch to establish sheet order
  reference_sheets <- openxlsx::getSheetNames(
    files[1]
  )
  analysis_sheets <- grep(
    analysis_sheet_pattern,
    reference_sheets,
    value = TRUE
  )
  if (length(analysis_sheets) == 0) {
    stop(
      "No MortaLIVE analysis sheets found."
    )
  }
  # Confirm all batches contain the same analysis sheets
  for (i in seq_along(files)) {
    current_sheets <- grep(
      analysis_sheet_pattern,
      openxlsx::getSheetNames(
        files[i]
      ),
      value = TRUE
    )
    if (
      !setequal(
        analysis_sheets,
        current_sheets
      )
    ) {
      stop(
        sprintf(
          "Analysis-sheet mismatch in %s.",
          basename(files[i])
        )
      )
    }
  }
  # Create a new clean workbook
  wb <- openxlsx::createWorkbook()
  # Preserve each source map separately when present.
  for (j in seq_along(files)) {
    if ("Well Map" %in% openxlsx::getSheetNames(files[j])) {
      map_sheet <- paste0("Well Map ", j)
      map <- openxlsx::read.xlsx(files[j], "Well Map", colNames = FALSE,
                                  skipEmptyRows = FALSE, skipEmptyCols = FALSE)
      openxlsx::addWorksheet(wb, map_sheet)
      openxlsx::writeData(wb, map_sheet, map, colNames = FALSE)
    }
  }
  manifest <- data.frame(Run_ID = batch_tags, Source_File = basename(files))
  openxlsx::addWorksheet(wb, "Sources")
  openxlsx::writeData(wb, "Sources", manifest)
  # Header formatting
  treatment_style <- openxlsx::createStyle(
    textDecoration = "bold",
    halign = "center",
    valign = "center"
  )
  well_style <- openxlsx::createStyle(
    halign = "center",
    valign = "center"
  )
  qc_rows <- vector(
    "list",
    length(analysis_sheets)
  )
  # Process every matching analysis sheet
  for (i in seq_along(analysis_sheets)) {
    sheet <- analysis_sheets[i]
    result <- combine_one_analysis_sheet(
      files = files,
      batch_tags = batch_tags,
      sheet = sheet
    )
    # Confirm values were copied unchanged
    verify_combined_sheet(
      result = result,
      sheet = sheet,
      batch_tags = batch_tags
    )
    openxlsx::addWorksheet(
      wb,
      sheet
    )
    # Row 1 = treatment headers
    openxlsx::writeData(
      wb,
      sheet,
      t(result$header1),
      startCol = 1,
      startRow = 1,
      colNames = FALSE,
      rowNames = FALSE,
      keepNA = FALSE
    )
    # Row 2 = batch-prefixed well IDs
    openxlsx::writeData(
      wb,
      sheet,
      t(result$header2),
      startCol = 1,
      startRow = 2,
      colNames = FALSE,
      rowNames = FALSE,
      keepNA = FALSE
    )
    # Row 3 onward = TIME + normalized values
    openxlsx::writeData(
      wb,
      sheet,
      result$data,
      startCol = 1,
      startRow = 3,
      colNames = FALSE,
      rowNames = FALSE,
      keepNA = FALSE
    )
    total_cols <- ncol(
      result$data
    )
    # Formatting
    openxlsx::addStyle(
      wb,
      sheet,
      treatment_style,
      rows = 1,
      cols = seq_len(total_cols),
      gridExpand = TRUE,
      stack = TRUE
    )
    openxlsx::addStyle(
      wb,
      sheet,
      well_style,
      rows = 2,
      cols = seq_len(total_cols),
      gridExpand = TRUE,
      stack = TRUE
    )
    openxlsx::setColWidths(
      wb,
      sheet,
      cols = 1,
      widths = 10
    )
    openxlsx::setColWidths(
      wb,
      sheet,
      cols = 2:total_cols,
      widths = 11
    )
    # Record QC information
    qc_rows[[i]] <- data.frame(
      sheet = sheet,
      timepoints =
        length(
          result$timepoints
        ),
      batches =
        length(files),
      source_slots_per_batch =
        paste(
          result$source_widths,
          collapse = " + "
        ),
      combined_slots_per_treatment =
        result$block_width,
      total_columns =
        total_cols,
      copy_qc =
        "PASS",
      stringsAsFactors = FALSE
    )
  }
  # Save final workbook
  openxlsx::saveWorkbook(
    wb,
    file = output_file,
    overwrite = TRUE
  )
  do.call(
    rbind,
    qc_rows
  )
}
```

```{r run-viability, eval=params$run_viability}
check_ids(names(viability_files), "viability_files names")
if (anyDuplicated(normalizePath(viability_files, mustWork = TRUE)))
  stop("A viability input file is listed more than once.")
normalized_files <- setNames(character(length(viability_files)), names(viability_files))
normalization_qc <- vector("list", length(viability_files))
for (i in seq_along(viability_files)) {
  id <- names(viability_files)[i]
  normalized_files[i] <- output_path("viability", paste0(id, "_VC100.xlsx"))
  normalization_qc[[i]] <- normalize_workbook_to_vc(viability_files[i], normalized_files[i])
  normalization_qc[[i]]$Run_ID <- id
}
normalization_qc <- do.call(rbind, normalization_qc)
combined_output <- output_path("viability", "combined_VC100.xlsx")
combination_qc <- combine_normalized_workbooks(
  normalized_files, names(normalized_files), combined_output)
stopifnot(all(normalization_qc$max_abs_error_from_100 <= QC_TOLERANCE),
          all(combination_qc$copy_qc == "PASS"))
openxlsx::write.xlsx(list(Normalization = normalization_qc, Combination = combination_qc),
                    output_path("viability", "quality_control.xlsx"))
print(normalization_qc)
print(combination_qc)
```

# Bliss independence mixture analysis — Figure S6

## Input and interpretation

The Excel sheet has a `dose` column and replicate columns such as `cd_1`, `cr_1`,
and `mix_1`. Measurements are percentages of the matched vehicle-control mean.
At each row, the single-agent doses must match the component doses in the mixture;
this code does not infer whether a reported mixture dose means total or per-component
concentration. Inputs must use the same endpoint and exposure time.

Fractional effect is $E=1-Y_{\%VC}/100$, and the expected effect is
$E_{Bliss}=E_{Cd}+E_{Cr}-E_{Cd}E_{Cr}$. Bliss excess is observed mixture effect
minus this expectation. Each treatment is bootstrapped independently within dose,
using percentile intervals. An interval spanning zero is labeled consistent with
Bliss additivity, not proof of equivalence. This procedure does not model experiment
clustering, matched samples, or uncertainty in the VC normalization denominator.
Values outside 0–100% are retained, not clipped; they trigger a warning because
fractional effects then fall outside the usual probability interpretation.

```{r bliss-functions}
read_bliss <- function(file, sheet) {
  check_file(file, sheet)
  raw <- janitor::clean_names(readxl::read_excel(file, sheet = sheet))
  if (!"dose" %in% names(raw)) stop("Bliss input requires a dose column.")
  sample_cols <- setdiff(names(raw), "dose")
  if (!length(sample_cols) || any(!grepl("^(cd|cr|mix)_[0-9]+$", sample_cols)))
    stop("Bliss replicate columns must be named cd_1, cr_1, mix_1, etc.")
  raw$dose <- strict_numeric(raw$dose, "Bliss dose")
  if (anyNA(raw$dose) || any(raw$dose < 0) || anyDuplicated(raw$dose))
    stop("Bliss doses must be unique, finite, nonnegative values.")
  raw[sample_cols] <- lapply(raw[sample_cols], strict_numeric, context = "Bliss measurements")
  long <- tidyr::pivot_longer(raw, cols = -dose,
    names_to = c("treatment", "replicate"), names_pattern = "^(cd|cr|mix)_([0-9]+)$",
    values_to = "percent_ld") %>%
    dplyr::mutate(replicate = as.integer(replicate), ld_ratio = percent_ld / 100,
                  effect = 1 - percent_ld / 100)
  if (any(long$percent_ld < 0 | long$percent_ld > 100, na.rm = TRUE))
    warning("Bliss input contains values outside 0–100%; no clipping was applied.")
  long
}
bootstrap_bliss <- function(data, n_boot = 5000L, seed = 831L) {
  if (n_boot < 2 || n_boot != as.integer(n_boot)) stop("n_boot must be an integer >= 2.")
  set.seed(seed)
  results <- lapply(unique(data$dose), function(current_dose) {
    d <- data[data$dose == current_dose, ]
    values <- lapply(c("cd", "cr", "mix"), function(treatment) {
      x <- d$effect[d$treatment == treatment]
      x <- x[is.finite(x)]
      if (length(x) < 2) stop("At least two finite replicates per treatment are required at dose ", current_dose)
      x
    })
    # Sampling indices avoids R's special interpretation of a length-one numeric vector.
    draw_mean <- function(x) mean(x[sample.int(length(x), length(x), replace = TRUE)])
    boot <- replicate(n_boot, {
      a <- draw_mean(values[[1]]); b <- draw_mean(values[[2]])
      ab <- draw_mean(values[[3]])
      ab - (a + b - a * b)
    })
    tibble::tibble(dose = current_dose, bootstrap_mean_excess = mean(boot),
                  ci_lower = unname(stats::quantile(boot, 0.025)),
                  ci_upper = unname(stats::quantile(boot, 0.975)))
  })
  dplyr::bind_rows(results)
}
```

```{r run-bliss, eval=params$run_bliss}
bliss_long <- read_bliss(bliss_input_file, bliss_input_sheet)
bliss_summary <- bliss_long %>%
  dplyr::group_by(dose, treatment) %>%
  dplyr::summarise(mean_effect = mean(effect, na.rm = TRUE),
    sd_effect = stats::sd(effect, na.rm = TRUE), n = sum(is.finite(effect)),
    n_missing = sum(!is.finite(effect)), .groups = "drop")
bliss_ci <- bootstrap_bliss(bliss_long, bliss_bootstrap_iterations, bliss_seed)
bliss_results <- bliss_summary %>%
  dplyr::select(dose, treatment, mean_effect) %>%
  tidyr::pivot_wider(names_from = treatment, values_from = mean_effect) %>%
  dplyr::mutate(bliss_expected = cd + cr - cd * cr,
                mix_observed = mix, bliss_excess = mix - bliss_expected) %>%
  dplyr::left_join(bliss_ci, by = "dose") %>%
  dplyr::mutate(interaction = dplyr::case_when(
    ci_upper < 0 ~ "Less than Bliss expectation",
    ci_lower > 0 ~ "Greater than Bliss expectation",
    TRUE ~ "Consistent with Bliss additivity"))
writexl::write_xlsx(list(Replicate_data = bliss_long, Treatment_summary = bliss_summary,
                         Bliss_results = bliss_results),
                    output_path("bliss", "Bliss_independence.xlsx"))
print(bliss_results)
```

# Sholl outgrowth analysis — Figures S7 and 4b–c

## Profiles, AUC, normalization, and testing

A formatted sheet contains `Radius` followed by replicate columns such as `VC_01`,
`0.32Cd_01`, `3.2Cr_02`, or `10Mix_03`. `Ctrl_01` and `Control_01` are also accepted
and standardized to `VC`. A manifest selects each sheet explicitly. Radius must
already represent the intended distance coordinate; this workflow does not identify
organoid edges or shift profiles. The default unit is µm, with no automatic pixel
conversion or subsampling. Set units and sampling explicitly for pixel-based inputs.

AUC is calculated for each profile by linear trapezoidal integration using
`MESS::auc(type = "linear")`. Missing observations are omitted; this connects the
remaining finite points across gaps and does not fill gaps with zero. The range and
number of points are exported. Profiles with fewer than two finite points are
excluded with an explicit QC status. Different radial ranges can affect AUC and
must be assessed when interpreting comparisons.

For each assay and source experiment, AUC is divided by the mean VC AUC. The
normalized VC mean is 1, not 100. Experiment-specific normalization is followed
by concatenation of replicate observations, without experiment-level averaging.
Two-sided, unpaired Wilcoxon rank-sum tests use `exact = FALSE` and the default
continuity correction. BH adjustment applies to **all dose-pair comparisons within
each metal**, not just the displayed VC comparisons. Tests do not include an
experiment/block effect. Shared controls are reused for each metal comparison,
not treated as additional independent observations in the exported replicate table.

```{r outgrowth-functions}
read_profile_table <- function(file, sheet, step = NULL, unit = "um") {
  check_file(file, sheet)
  raw <- readxl::read_excel(file, sheet = sheet, .name_repair = "minimal")
  if (!"Radius" %in% names(raw) || ncol(raw) < 2 || anyDuplicated(names(raw)))
    stop("A profile sheet requires one Radius column and unique sample columns: ", file)
  raw$Radius <- strict_numeric(raw$Radius, paste(file, "Radius"))
  raw[setdiff(names(raw), "Radius")] <- lapply(raw[setdiff(names(raw), "Radius")],
      strict_numeric, context = paste(file, "intersections"))
  if (any(is.na(raw$Radius) & rowSums(!is.na(raw[setdiff(names(raw), "Radius")])) > 0))
    stop("Intersections have no radius in ", file)
  raw <- raw[!is.na(raw$Radius), ]
  if (anyDuplicated(raw$Radius) || any(raw$Radius < 0))
    stop("Radii must be unique and nonnegative in ", file)
  if (!is.null(step)) {
    if (!is.finite(step) || step <= 0) stop("Sampling step must be positive.")
    raw <- raw[abs(raw$Radius / step - round(raw$Radius / step)) < 1e-8, ]
  }
  if (nrow(raw) < 2) stop("Fewer than two radius rows remain in ", file)
  long <- tidyr::pivot_longer(raw, cols = -Radius, names_to = "Sample", values_to = "Inters")
  if (any(long$Inters < 0, na.rm = TRUE)) stop("Negative intersection counts in ", file)
  long$Radius_unit <- unit
  long
}
parse_metal_samples <- function(long, metals = outgrowth_metals) {
  sample <- trimws(long$Sample)
  ctrl <- grepl("^(VC|Ctrl|Control)_?[0-9]+$", sample, ignore.case = TRUE)
  parsed <- stringr::str_match(sample, "^([0-9]+(?:\\.[0-9]+)?)([A-Za-z]+|Cd\\+Cr)_?([0-9]+)$")
  metal <- ifelse(ctrl, "VC", parsed[, 3])
  metal[metal == "Cd+Cr"] <- "Mix"
  dose <- suppressWarnings(as.numeric(parsed[, 2])); dose[ctrl] <- 0
  rep <- ifelse(ctrl, stringr::str_extract(sample, "[0-9]+$"), parsed[, 4])
  bad <- is.na(dose) | is.na(metal) | is.na(rep) | !metal %in% c("VC", metals)
  if (any(bad)) stop("Unrecognized sample names: ", paste(unique(sample[bad]), collapse = ", "))
  if (any(!ctrl & dose <= 0)) stop("Non-VC treatments must have positive doses.")
  long %>% dplyr::mutate(Sample = sample, Metal = metal, Dose_uM = dose, Rep = rep)
}
profile_auc <- function(long, groups) {
  long %>% dplyr::group_by(dplyr::across(dplyr::all_of(groups))) %>%
    dplyr::group_modify(function(dat, key) {
      ok <- is.finite(dat$Radius) & is.finite(dat$Inters)
      x <- dat$Radius[ok]; y <- dat$Inters[ok]; ord <- order(x)
      if (anyDuplicated(x)) stop("A replicate contains duplicated radii; check source identifiers.")
      n <- length(x)
      tibble::tibble(AUC = if (n >= 2) MESS::auc(x[ord], y[ord], type = "linear") else NA_real_,
        n_points = n, n_missing = sum(!ok),
        x_min = if (n) min(x) else NA_real_, x_max = if (n) max(x) else NA_real_,
        Status = if (n >= 2) "Included" else "Excluded: fewer than two finite points")
    }) %>% dplyr::ungroup()
}
control_by_metal <- function(dat) {
  metals <- unique(dat$Metal[dat$Metal != "VC"])
  vc <- dat %>% dplyr::filter(Metal == "VC", Dose_uM == 0) %>% dplyr::select(-Metal)
  if (!nrow(vc)) stop("No vehicle controls available.")
  dplyr::bind_rows(dplyr::filter(dat, Metal != "VC"), tidyr::crossing(vc, Metal = metals))
}
wilcox_result <- function(x, y, exact = FALSE, minimum_n = 1L) {
  x <- x[is.finite(x)]; y <- y[is.finite(y)]
  warnings <- character()
  test <- if (length(x) >= minimum_n && length(y) >= minimum_n) {
    tryCatch(withCallingHandlers(stats::wilcox.test(x, y, alternative = "two.sided",
      paired = FALSE, exact = exact, correct = TRUE), warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w)); invokeRestart("muffleWarning")
      }), error = function(e) { warnings <<- conditionMessage(e); NULL })
  } else NULL
  p <- if (is.null(test)) NA_real_ else test$p.value
  tibble::tibble(n1 = length(x), n2 = length(y),
    median1 = if (length(x)) stats::median(x) else NA_real_,
    median2 = if (length(y)) stats::median(y) else NA_real_,
    W = if (is.null(test)) NA_real_ else unname(test$statistic), p = p,
    low_n_flag = length(x) < 3 || length(y) < 3,
    Test_Status = if (is.finite(p)) "Tested" else "Not estimable or insufficient n",
    Test_Warning = paste(warnings, collapse = "; "))
}
pairwise_metals <- function(dat, value, minimum_n, exact = FALSE) {
  purrr::map_dfr(sort(unique(dat$Metal)), function(m) {
    d <- dat[dat$Metal == m, ]; doses <- sort(unique(d$Dose_uM))
    if (length(doses) < 2) stop("At least two doses including VC are required for ", m)
    pairs <- utils::combn(doses, 2)
    purrr::map_dfr(seq_len(ncol(pairs)), function(i) {
      a <- pairs[1, i]; b <- pairs[2, i]
      dplyr::bind_cols(tibble::tibble(Metal = m, dose1 = a, dose2 = b),
        wilcox_result(d[[value]][d$Dose_uM == a], d[[value]][d$Dose_uM == b], exact, minimum_n))
    }) %>% dplyr::mutate(p_adj = stats::p.adjust(p, method = wilcoxon_adjustment))
  })
}
auc_summary <- function(dat, value, groups) {
  dat %>% dplyr::filter(is.finite(.data[[value]])) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(groups))) %>%
    dplyr::summarise(n = dplyr::n(), mean = mean(.data[[value]]),
      SD = stats::sd(.data[[value]]), SEM = SD / sqrt(n),
      median = stats::median(.data[[value]]),
      CV_percent = if (mean != 0) 100 * SD / mean else NA_real_, .groups = "drop")
}
normality_summary <- function(dat, value, groups) {
  dat %>% dplyr::group_by(dplyr::across(dplyr::all_of(groups))) %>%
    dplyr::group_modify(function(d, key) {
      x <- d[[value]][is.finite(d[[value]])]
      valid <- length(x) >= 3 && length(x) <= 5000 && length(unique(x)) > 1
      tibble::tibble(n = length(x), shapiro_p = if (valid) stats::shapiro.test(x)$p.value else NA_real_,
                    Status = if (valid) "Tested" else "Requires 3–5000 nonconstant observations")
    }) %>% dplyr::ungroup()
}
plot_metal_auc <- function(dat, tests, value, title, cutoff, normalized = FALSE) {
  stopifnot(plot_error %in% c("SD", "SEM"))
  d <- control_by_metal(dat) %>% dplyr::filter(is.finite(.data[[value]])) %>%
    dplyr::mutate(Dose_f = factor(Dose_uM, levels = outgrowth_doses_uM))
  s <- auc_summary(d, value, c("Metal", "Dose_f")); s$Error <- s[[plot_error]]
  p <- ggplot2::ggplot(d, ggplot2::aes(Dose_f, .data[[value]], fill = Dose_f)) +
    ggplot2::geom_col(data = s, ggplot2::aes(y = mean), width = 0.65) +
    ggplot2::geom_errorbar(data = s, ggplot2::aes(y = mean, ymin = mean - Error, ymax = mean + Error), width = 0.2) +
    ggplot2::geom_point(shape = 21, position = ggplot2::position_jitter(width = 0.08, height = 0, seed = plot_seed)) +
    ggplot2::facet_wrap(~Metal, nrow = 1) + ggplot2::theme_classic() +
    ggplot2::theme(legend.position = "none") +
    ggplot2::labs(title = title, subtitle = paste("Mean ±", plot_error),
      x = "Exposure (µM)", y = if (normalized) "AUC / experiment VC mean" else "AUC")
  if (normalized) p <- p + ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed")
  ann <- tests %>% dplyr::filter(dose1 == 0, is.finite(p_adj), p_adj < cutoff) %>%
    dplyr::mutate(group1 = as.character(dose1), group2 = as.character(dose2),
      label = ifelse(p_adj < 0.001, "<0.001", sprintf("%.3f", p_adj)))
  if (nrow(ann)) {
    ymax <- d %>% dplyr::group_by(Metal) %>%
      dplyr::summarise(y0 = max(.data[[value]], na.rm = TRUE), .groups = "drop")
    ann <- ann %>% dplyr::left_join(ymax, by = "Metal") %>% dplyr::group_by(Metal) %>%
      dplyr::mutate(y.position = pmax(y0, 1e-6) * (1.1 + 0.08 * (dplyr::row_number() - 1))) %>%
      dplyr::ungroup()
    p <- p + ggpubr::stat_pvalue_manual(ann, label = "label", xmin = "group1", xmax = "group2",
                y.position = "y.position", inherit.aes = FALSE, tip.length = 0.01)
  }
  p
}
```

## Shared profile summaries and comparison plots

```{r profile-comparison-helpers}
read_profile_manifest <- function(manifest, step, unit, scale = 1) {
  needed <- c("Source_ID", "File", "Sheet")
  if (!all(needed %in% names(manifest)) || !nrow(manifest)) stop("A profile manifest requires Source_ID, File and Sheet.")
  check_ids(manifest$Source_ID, "Profile source IDs")
  if (anyDuplicated(paste(normalizePath(manifest$File, mustWork = TRUE), manifest$Sheet)))
    stop("The same file/sheet is listed more than once.")
  if (!is.finite(scale) || scale <= 0) stop("Radius scale must be positive.")
  purrr::map_dfr(seq_len(nrow(manifest)), function(i) {
    row <- manifest[i, ]
    d <- read_profile_table(row$File, row$Sheet, step, unit)
    d$Radius <- d$Radius * scale
    d$Source_ID <- row$Source_ID
    d$Sample_UID <- paste(row$Source_ID, d$Sample, sep = "__")
    for (name in setdiff(names(manifest), c("File", "Sheet", "Source_ID"))) d[[name]] <- row[[name]]
    d
  })
}
profile_summary <- function(long, group) {
  long %>% dplyr::filter(is.finite(Inters)) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(group)), Radius) %>%
    dplyr::summarise(n = dplyr::n(), mean_intersections = mean(Inters),
      SD = stats::sd(Inters), SEM = SD / sqrt(n),
      CI95_half_width = if (n > 1) stats::qt(0.975, n - 1) * SEM else NA_real_,
      .groups = "drop")
}
plot_profiles <- function(summary, group, title, unit, ribbon = "SEM") {
  stopifnot(ribbon %in% c("SEM", "SD", "CI95_half_width"))
  s <- summary; s$Error <- s[[ribbon]]
  ggplot2::ggplot(s, ggplot2::aes(Radius, mean_intersections,
      color = .data[[group]], fill = .data[[group]])) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = mean_intersections - Error,
      ymax = mean_intersections + Error), alpha = 0.18, color = NA) +
    ggplot2::geom_line(linewidth = 0.8) + ggplot2::theme_classic() +
    ggplot2::labs(title = title, subtitle = paste("Mean with", ribbon, "ribbon"),
      x = paste0("Distance from organoid edge (", unit, ")"), y = "Intersections", color = NULL, fill = NULL)
}
plot_group_auc <- function(dat, group, value, tests, title, normalized = FALSE) {
  p <- ggplot2::ggplot(dat, ggplot2::aes(.data[[group]], .data[[value]], fill = .data[[group]])) +
    ggplot2::geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.65) +
    ggplot2::geom_point(shape = 21, position = ggplot2::position_jitter(width = 0.12, height = 0, seed = plot_seed)) +
    ggplot2::theme_classic() + ggplot2::theme(legend.position = "none") +
    ggplot2::labs(title = title, x = NULL, y = if (normalized) "AUC / control mean" else "AUC")
  if (normalized) p <- p + ggplot2::geom_hline(yintercept = 1, linetype = "dashed")
  ann <- tests[is.finite(tests$p), ]
  if (nrow(ann)) {
    y0 <- max(dat[[value]], na.rm = TRUE)
    ann$y.position <- pmax(y0, 1e-6) * (1.08 + 0.08 * seq_len(nrow(ann)))
    ann$label <- ifelse(ann$p < 0.001, "p < 0.001", paste0("p = ", sprintf("%.3f", ann$p)))
    p <- p + ggpubr::stat_pvalue_manual(ann, label = "label", xmin = "group1", xmax = "group2",
      y.position = "y.position", inherit.aes = FALSE, tip.length = 0.01)
  }
  p
}
```

```{r run-outgrowth, eval=params$run_outgrowth}
required_cols <- c("Assay", "Run_ID", "File", "Sheet")
if (!all(required_cols %in% names(outgrowth_inputs)) || !nrow(outgrowth_inputs))
  stop("outgrowth_inputs requires Assay, Run_ID, File and Sheet.")
check_ids(unique(outgrowth_inputs$Assay), "Assay IDs")
for (assay in unique(outgrowth_inputs$Assay))
  check_ids(outgrowth_inputs$Run_ID[outgrowth_inputs$Assay == assay], "Run IDs within each assay")
if (anyDuplicated(paste(normalizePath(outgrowth_inputs$File, mustWork = TRUE), outgrowth_inputs$Sheet)))
  stop("The same outgrowth file/sheet is listed twice.")
all_auc <- vector("list", nrow(outgrowth_inputs))
for (i in seq_len(nrow(outgrowth_inputs))) {
  spec <- outgrowth_inputs[i, ]; prefix <- paste(spec$Assay, spec$Run_ID, sep = "_")
  long <- read_profile_table(spec$File, spec$Sheet, outgrowth_radius_step, outgrowth_radius_unit) %>%
    parse_metal_samples() %>% dplyr::filter(Dose_uM %in% outgrowth_doses_uM) %>%
    dplyr::mutate(Assay = spec$Assay, Run_ID = spec$Run_ID, Source_File = basename(spec$File))
  auc <- profile_auc(long, c("Assay", "Run_ID", "Source_File", "Sample", "Metal", "Dose_uM", "Rep"))
  vc <- auc$AUC[auc$Metal == "VC" & is.finite(auc$AUC)]
  if (!length(vc) || mean(vc) <= 0) stop("Missing or nonpositive VC AUC for ", prefix)
  auc <- auc %>% dplyr::mutate(Raw_AUC = AUC, VC_mean_AUC = mean(vc),
    Normalized_AUC = AUC / mean(vc), Sample_UID = paste(Assay, Run_ID, Sample, sep = "__"))
  tests <- pairwise_metals(control_by_metal(auc), "AUC", single_run_min_n, outgrowth_exact)
  raw_summary <- auc_summary(auc, "AUC", c("Metal", "Dose_uM"))
  norm_summary <- auc_summary(auc, "Normalized_AUC", c("Metal", "Dose_uM"))
  normality <- normality_summary(auc, "AUC", c("Metal", "Dose_uM"))
  curve_summary <- profile_summary(control_by_metal(long), c("Metal", "Dose_uM"))
  curve_summary$Dose_label <- factor(curve_summary$Dose_uM, levels = outgrowth_doses_uM)
  p_curve <- plot_profiles(curve_summary, "Dose_label", prefix, outgrowth_radius_unit) +
    ggplot2::facet_wrap(~Metal, nrow = 1, scales = "free_y")
  ggplot2::ggsave(output_path("outgrowth", paste0(prefix, "_intersection_traces.png")),
                 p_curve, width = 12, height = 4, dpi = 300)
  print(p_curve)
  writexl::write_xlsx(list(Long_profiles = long, Normalized_AUC_per_replicate = auc,
    Raw_summary = raw_summary, Normalized_summary = norm_summary, Curve_summary = curve_summary,
    Pairwise_Wilcoxon = tests, Normality = normality),
    output_path("outgrowth", paste0(prefix, "_AUC.xlsx")))
  readr::write_csv(auc, output_path("outgrowth", paste0(prefix, "_normalized_AUC_per_replicate.csv")))
  for (metric in c("AUC", "Normalized_AUC")) {
    p <- plot_metal_auc(auc, tests, metric, prefix, single_run_display_cutoff, metric == "Normalized_AUC")
    ggplot2::ggsave(output_path("outgrowth", paste0(prefix, "_", metric, ".png")), p, width = 12, height = 5, dpi = 300)
    print(p)
  }
  all_auc[[i]] <- auc
}
outgrowth_all <- dplyr::bind_rows(all_auc)
for (assay in unique(outgrowth_all$Assay)) {
  d <- outgrowth_all %>% dplyr::filter(Assay == assay, is.finite(Normalized_AUC))
  qc <- d %>% dplyr::filter(Metal == "VC") %>% dplyr::group_by(Run_ID) %>%
    dplyr::summarise(n_VC = dplyr::n(), mean_VC = mean(Normalized_AUC), .groups = "drop")
  stopifnot(all(abs(qc$mean_VC - 1) <= QC_TOLERANCE), !anyDuplicated(d$Sample_UID))
  tests <- pairwise_metals(control_by_metal(d), "Normalized_AUC", pooled_min_n, FALSE)
  summary <- auc_summary(d, "Normalized_AUC", c("Metal", "Dose_uM"))
  sources <- d %>% dplyr::group_by(Metal, Dose_uM) %>%
    dplyr::summarise(n_sources = dplyr::n_distinct(Run_ID), .groups = "drop")
  summary <- dplyr::left_join(summary, sources, by = c("Metal", "Dose_uM"))
  writexl::write_xlsx(list(Combined_per_replicate = d, Pooled_summary = summary,
    Pairwise_Wilcoxon = tests, VC_comparisons = dplyr::filter(tests, dose1 == 0), Source_QC = qc),
    output_path("outgrowth", paste0(assay, "_combined_AUC.xlsx")))
  readr::write_csv(d, output_path("outgrowth", paste0(assay, "_combined_normalized_AUC_per_replicate.csv")))
  p <- plot_metal_auc(d, tests, "Normalized_AUC", paste(assay, "combined"), pooled_display_cutoff, TRUE)
  ggplot2::ggsave(output_path("outgrowth", paste0(assay, "_combined_AUC.png")), p, width = 12, height = 5, dpi = 300)
  print(p)
}
```

# Outgrowth concentration–response fitting — Figure 4d

The model fixes the upper and lower asymptotes at 100 and 0, respectively:

\[
Y=\frac{100}{1+10^{h(\log_{10}IC_{50}-\log_{10}x)}}.
\]

The Hill slope is constrained to [-100, 0] (negative slopes give decreasing
responses in this parameterization), and log10(IC50) to [-12, log10(1000)]. Vehicle
predictions are fixed at 100; VC observations still contribute to residual statistics.
Multiple starting values are fitted with `minpack.lm::nlsLM`; the successful candidate
with the lowest residual sum of squares is retained. Inputs are replicate-level
normalized AUC **ratios**, multiplied by 100 internally. ML and NO are fitted separately.

The active source workflow used `confint()` for nonlinear-model profile intervals;
its separately defined extra-sum-of-squares profiling helper was never called.
Only the active method is retained. Raw interval estimates and their status are
exported. The source's capped upper-limit column is also retained, but a failed
interval is explicitly labeled unestimated rather than evidence that IC50 exceeds
the cap. An extrapolated or boundary-limited IC50 requires cautious interpretation.

```{r ic50-model-functions}
# Prism-style normalized inhibitor-response model
#
# Top = 100
# Bottom = 0
#
# Y = 100 / [1 + (IC50/X)^HillSlope]
dose_response <- function(
  x,
  logIC50,
  HillSlope
) {
  # Avoid log10(0)
  x_safe <- pmax(
    x,
    1e-12
  )
  y <- 100 / (
    1 +
      10^(
        HillSlope *
          (
            logIC50 -
              log10(x_safe)
          )
      )
  )
  # Vehicle control
  y[x == 0] <- 100
  y
}
# -----------------------------------------------------------------------------
# Fit one dose-response curve
# -----------------------------------------------------------------------------
fit_ic50 <- function(dat) {
  # Try multiple starting IC50 and HillSlope values.
  # Multiple starts reduce sensitivity to the initial parameter guess.
  starts <- expand.grid(
    logIC50 = log10(
      c(
        0.32,
        1,
        3.2,
        10,
        30,
        100,
        300
      )
    ),
    HillSlope = c(
      -0.5,
      -1,
      -2,
      -5,
      -10,
      -20
    )
  )
  candidate_fits <- lapply(
    seq_len(nrow(starts)),
    function(i) {
      tryCatch(
        nlsLM(
          Response ~
            dose_response(
              Dose_uM,
              logIC50,
              HillSlope
            ),
          data = dat,
          start = list(
            logIC50 =
              starts$logIC50[i],
            HillSlope =
              starts$HillSlope[i]
          ),
          lower = c(
            logIC50 = -12,
            HillSlope = -100
          ),
          upper = c(
            logIC50 = max_logIC50,
            HillSlope = 0
          ),
          control = nls.lm.control(
            maxiter = max_iterations
          )
        ),
        error = function(e) {
          NULL
        }
      )
    }
  )
  # Remove failed starting points
  candidate_fits <- Filter(function(x) !is.null(x) && isTRUE(x$convInfo$isConv), candidate_fits)
  if (length(candidate_fits) == 0) {
    stop(
      "No valid nonlinear fit could be obtained."
    )
  }
  # Calculate residual sum of squares
  RSS <- sapply(
    candidate_fits,
    function(fit) {
      sum(
        residuals(fit)^2
      )
    }
  )
  # Keep the best-fitting solution
  candidate_fits[[which.min(RSS)]]
}
```

```{r ic50-results-function}
get_ic50_results <- function(fit) {
  # Best-fit parameters
  pars <- coef(fit)
  logIC50 <- unname(pars["logIC50"])
  HillSlope <- unname(pars["HillSlope"])
  IC50 <- 10^logIC50
  # ---------------------------------------------------------------------------
  # 95% profile-likelihood confidence intervals
  # ---------------------------------------------------------------------------
  CI <- tryCatch(
    suppressMessages(
      confint(
        fit,
        level = 0.95
      )
    ),
    error = function(e) {
      matrix(
        NA_real_,
        nrow = 2,
        ncol = 2,
        dimnames = list(
          c("logIC50", "HillSlope"),
          c("2.5%", "97.5%")
        )
      )
    }
  )
  # ---------------------------------------------------------------------------
  # IC50 confidence interval
  # ---------------------------------------------------------------------------
  logIC50_lower <- CI["logIC50", 1]
  logIC50_upper <- CI["logIC50", 2]
  IC50_lower <- 10^logIC50_lower
  IC50_upper <- 10^logIC50_upper
  # Preserve unmodified estimates separately from the source-style display cap.
  raw_lower <- IC50_lower
  raw_upper <- IC50_upper
  ci_status <- if (is.na(raw_lower) || is.na(raw_upper)) "Not fully estimated" else
    if (!is.finite(raw_lower) || !is.finite(raw_upper)) "Unbounded" else "Estimated"
  # A capped or unresolved interval is not a finite confidence bound.
  upper_capped <- (
    is.na(IC50_upper) ||
    is.infinite(IC50_upper) ||
    IC50_upper >= max_IC50
  )
  if (upper_capped) {
    IC50_upper <- max_IC50
    logIC50_upper <- max_logIC50
  }
  IC50_CI_text <- if (ci_status == "Not fully estimated") {
    "Not fully estimated; inspect raw bounds"
  } else paste0(signif(raw_lower, 4), " to ",
    if (upper_capped) paste0(">=", max_IC50, " (reporting cap)") else signif(raw_upper, 4))
  # ---------------------------------------------------------------------------
  # Goodness of fit
  # ---------------------------------------------------------------------------
  residuals_fit <- residuals(fit)
  fitted_values <- fitted(fit)
  # Reconstruct observed Y values
  observed <- fitted_values + residuals_fit
  RSS <- sum(
    residuals_fit^2
  )
  TSS <- sum(
    (observed - mean(observed))^2
  )
  df <- df.residual(fit)
  R_squared <- if (TSS > 0) 1 - RSS / TSS else NA_real_
  Sy_x <- sqrt(
    RSS / df
  )
  # ---------------------------------------------------------------------------
  # Return results
  # ---------------------------------------------------------------------------
  tibble(
    IC50_uM = IC50,
    Raw_IC50_lower_95_uM = raw_lower,
    Raw_IC50_upper_95_uM = raw_upper,
    CI_Status = ci_status,
    Parameter_boundary = abs(logIC50 - max_logIC50) < 1e-6 || abs(logIC50 + 12) < 1e-6 ||
                         abs(HillSlope) < 1e-6 || abs(HillSlope + 100) < 1e-6,
    IC50_lower_95_uM = IC50_lower,
    IC50_upper_95_uM = IC50_upper,
    IC50_95_CI = IC50_CI_text,
    HillSlope = HillSlope,
    HillSlope_lower_95 = CI["HillSlope", 1],
    HillSlope_upper_95 = CI["HillSlope", 2],
    logIC50 = logIC50,
    logIC50_lower_95 = logIC50_lower,
    logIC50_upper_95 = logIC50_upper,
    Degrees_of_Freedom = df,
    R_squared = R_squared,
    Sum_of_Squares = RSS,
    Sy_x = Sy_x,
    Number_of_points = length(observed),
    IC50_upper_capped = upper_capped
  )
}
```

```{r run-ic50, eval=params$run_ic50}
check_ids(names(ic50_files), "ic50_files assay names")
ic50_data <- purrr::imap_dfr(ic50_files, function(path, assay) {
  check_file(path)
  d <- readr::read_csv(path, show_col_types = FALSE)
  if (!all(c("Metal", "Dose_uM", "Normalized_AUC") %in% names(d)))
    stop("IC50 inputs require Metal, Dose_uM and Normalized_AUC: ", path)
  d$Dose_uM <- strict_numeric(d$Dose_uM, "IC50 dose")
  d$Normalized_AUC <- strict_numeric(d$Normalized_AUC, "IC50 response")
  if (any(d$Dose_uM < 0, na.rm = TRUE)) stop("IC50 doses cannot be negative.")
  d %>% dplyr::mutate(Assay_ID = assay, Response = Normalized_AUC * 100)
})
fits <- list()
ic50_results <- purrr::map_dfr(names(ic50_files), function(assay) {
  purrr::map_dfr(ic50_metals, function(metal) {
    d <- ic50_data %>% dplyr::filter(Assay_ID == assay, Metal %in% c("VC", metal),
                                    is.finite(Dose_uM), is.finite(Response))
    result <- tryCatch({
      if (nrow(d) <= 2 || length(unique(d$Dose_uM[d$Dose_uM > 0])) < 2)
        stop("At least two distinct positive doses and >2 observations required.")
      fit <- fit_ic50(d)
      fits[[paste(assay, metal, sep = "_")]] <<- fit
      get_ic50_results(fit) %>% dplyr::mutate(Fit_Status = "Fitted", Fit_Message = "")
    }, error = function(e) tibble::tibble(Fit_Status = "Failed", Fit_Message = conditionMessage(e)))
    dplyr::bind_cols(tibble::tibble(Assay = assay, Metal = metal), result)
  })
})
writexl::write_xlsx(list(IC50_results = ic50_results), output_path("ic50", "Outgrowth_IC50.xlsx"))
saveRDS(fits, output_path("ic50", "Outgrowth_IC50_fits.rds"))
print(ic50_results)
```

# MortaLIVE compatibility validation — Figure 1d–e

The β-III-tubulin and GFAP datasets are processed separately. Each sample is assigned
to an unstained control or a MortaLIVE dilution from its column name. AUC is integrated
in the configured radius units, then divided by the pooled control mean **within
that marker dataset**, following the validation scripts. This normalization differs
from the experiment-specific normalization used for exposure analyses above.

Each dilution is compared with control using a two-sided, unpaired Wilcoxon rank-sum
test. Normalized-AUC tests use `exact = FALSE`, with no multiplicity adjustment, as
in both validation scripts. For provenance, raw-AUC tests retain the GFAP script's
`exact = TRUE` request for the lower dilution and `FALSE` for the other comparisons;
R may fall back to an approximation in the presence of ties. All warnings are exported.
Changing exact-test settings can change p-values even though positive scaling
preserves ranks. Box plots show the median and IQR, with whiskers to the most extreme
observations within 1.5 IQR; points display the individual observations.



```{r run-validation, eval=params$run_validation}
if (!"Marker" %in% names(validation_inputs)) stop("validation_inputs requires a Marker column.")
check_ids(unique(validation_inputs$Marker), "Marker IDs")
validation_long <- read_profile_manifest(validation_inputs, validation_radius_step,
                                         validation_radius_unit, validation_radius_scale)
validation_long <- validation_long %>% dplyr::mutate(
  Group = dplyr::case_when(
    grepl("^(ctrl|control|VC)", Sample, ignore.case = TRUE) ~ "Ctrl",
    TRUE ~ sub("_[0-9]+$", "", Sample)))
if (any(!validation_long$Group %in% validation_groups))
  stop("Unrecognized validation sample(s): ", paste(unique(validation_long$Sample[!validation_long$Group %in% validation_groups]), collapse = ", "))
for (marker in unique(validation_long$Marker)) {
  d <- validation_long[validation_long$Marker == marker, ]
  d$Group <- factor(d$Group, levels = validation_groups)
  auc <- profile_auc(d, c("Marker", "Source_ID", "Sample", "Sample_UID", "Group"))
  ctrl <- auc$AUC[auc$Group == "Ctrl" & is.finite(auc$AUC)]
  if (!length(ctrl) || mean(ctrl) <= 0) stop("Invalid control AUC for ", marker)
  auc$AUC_norm <- auc$AUC / mean(ctrl)
  # Zero AUC gives log2FC = -Inf; no pseudocount is introduced.
  auc$log2FC <- log2(auc$AUC_norm)
  tables <- lapply(c("AUC", "AUC_norm"), function(value) {
    purrr::map_dfr(setdiff(validation_groups, "Ctrl"), function(group) {
      exact <- value == "AUC" && marker == "GFAP" && group == "0.0005ML"
      dplyr::bind_cols(tibble::tibble(group1 = "Ctrl", group2 = group,
                                    Metric = value, Exact_requested = exact),
        wilcox_result(auc[[value]][auc$Group == "Ctrl"], auc[[value]][auc$Group == group], exact))
    })
  })
  tests <- dplyr::bind_rows(tables)
  summary <- profile_summary(d, "Group")
  norm_summary <- auc_summary(auc, "AUC_norm", "Group")
  writexl::write_xlsx(list(Long_profiles = d, AUC_per_replicate = auc,
    Normalized_summary = norm_summary, Profile_summary = summary, Wilcoxon = tests),
    output_path("validation", paste0(marker, "_validation.xlsx")))
  p_curve <- plot_profiles(summary, "Group", paste(marker, "validation"), validation_radius_unit)
  p_raw <- plot_group_auc(auc, "Group", "AUC", tests[tests$Metric == "AUC", ], paste(marker, "raw AUC"))
  p_norm <- plot_group_auc(auc, "Group", "AUC_norm", tests[tests$Metric == "AUC_norm", ],
                           paste(marker, "normalized AUC"), TRUE)
  for (name in c("curve", "raw", "normalized")) {
    p <- switch(name, curve = p_curve, raw = p_raw, normalized = p_norm)
    ggplot2::ggsave(output_path("validation", paste0(marker, "_", name, ".png")), p, width = 8, height = 5, dpi = 300)
    print(p)
  }
}
```

# MortaLIVE versus immunofluorescence — Figure 3c–d

A manifest assigns each workbook/sheet to a measurement method; filenames do not
need to encode method. The same function can reproduce the two-method comparison
or the β-III-tubulin/GFAP/ML comparison. An additional combined-marker input may
be included if measured profiles are available. Combined-marker intersections
are **not calculated by adding separate marker counts**, which can double-count
spatially overlapping intersections.

Profile plots show mean intersections with SEM by default, matching the scripts.
`comparison_ribbon = "CI95_half_width"` explicitly requests a pointwise Student-t
95% confidence interval; this is an option, not the original script's calculation.
The summary exports both quantities and the peak radius of the mean profile.

AUC comparisons use unpaired, two-sided Wilcoxon rank-sum tests (`exact = FALSE`),
with no multiplicity correction. Each configured comparator is tested against ML.
Even when sample suffixes identify matched organoids, the original scripts used
unpaired tests. This workflow retains that choice and does not silently substitute
a paired analysis. Biological pairing must be considered separately when interpreting
or revising the statistical plan.

```{r run-method-comparison, eval=params$run_comparison}
if (!"Method" %in% names(comparison_inputs)) stop("comparison_inputs requires a Method column (NA permits name parsing).")
comparison_long <- read_profile_manifest(comparison_inputs, comparison_radius_step,
                                         comparison_radius_unit, comparison_radius_scale)
# Method may be omitted only for columns carrying an explicit _IF_ or _ML_ token.
missing_method <- is.na(comparison_long$Method) | comparison_long$Method == ""
parsed_method <- ifelse(grepl("_IF_", comparison_long$Sample), "IF",
                  ifelse(grepl("_ML_", comparison_long$Sample), "ML", NA_character_))
comparison_long$Method[missing_method] <- parsed_method[missing_method]
if (anyNA(comparison_long$Method)) stop("Specify a Method for every comparison input.")
if (!comparison_reference %in% comparison_long$Method) stop("Reference method not present.")
comparison_long$Method <- factor(comparison_long$Method,
  levels = c(setdiff(unique(comparison_long$Method), comparison_reference), comparison_reference))
comparison_auc <- profile_auc(comparison_long, c("Source_ID", "Sample", "Sample_UID", "Method"))
comparison_summary <- profile_summary(comparison_long, "Method")
comparison_peaks <- comparison_summary %>% dplyr::group_by(Method) %>%
  dplyr::filter(mean_intersections == max(mean_intersections)) %>% dplyr::ungroup()
comparators <- setdiff(levels(comparison_long$Method), comparison_reference)
if (!length(comparators)) stop("At least one comparator method is required.")
comparison_tests <- purrr::map_dfr(comparators, function(method) {
  dplyr::bind_cols(tibble::tibble(group1 = method, group2 = comparison_reference),
    wilcox_result(comparison_auc$AUC[comparison_auc$Method == method],
                  comparison_auc$AUC[comparison_auc$Method == comparison_reference], FALSE))
})
comparison_auc_summary <- auc_summary(comparison_auc, "AUC", "Method")
writexl::write_xlsx(list(Long_profiles = comparison_long, Profile_summary = comparison_summary,
  Mean_profile_peaks = comparison_peaks, AUC_per_replicate = comparison_auc,
  AUC_summary = comparison_auc_summary, Wilcoxon = comparison_tests),
  output_path("method_comparison", "ML_IF_comparison.xlsx"))
p_trace <- plot_profiles(comparison_summary, "Method", "Outgrowth profiles", comparison_radius_unit, comparison_ribbon)
p_auc <- plot_group_auc(comparison_auc, "Method", "AUC", comparison_tests, "MortaLIVE and immunofluorescence AUC")
ggplot2::ggsave(output_path("method_comparison", "intersection_traces.png"), p_trace, width = 9, height = 5, dpi = 300)
ggplot2::ggsave(output_path("method_comparison", "raw_AUC_comparison.png"), p_auc, width = 8, height = 5, dpi = 300)
print(p_trace)
print(p_auc)
```

# Optional Fiji Sholl CSV assembly

This utility combines `*_Sholl-Profiles.csv` files within each immediate subfolder
of a configured parent directory. CSVs must contain `Radius` and `Inters.`. Profiles
are joined on the actual radius values, preserving missing cells. Distances are not
rebased to the organoid edge. Duplicate sample names or radii stop the import.

Output columns retain file-derived sample IDs. These IDs must be mapped to the
appropriate exposure or validation labels before running the figure analyses.
This utility does not implement a plate map or infer experimental conditions.

```{r sholl-assembly-functions}
extract_sample_name <- function(file_path, folder_name) {
  name <- sub("\\.tif_Sholl-Profiles\\.csv$", "", basename(file_path))
  name <- sub("_Sholl-Profiles\\.csv$", "", name)
  if (startsWith(name, folder_name)) name <- substring(name, nchar(folder_name) + 1L)
  well <- stringr::str_extract(name, "[A-Z]-[0-9]+_F[0-9]+")
  if (is.na(well)) name else well
}
combine_sholl_folder <- function(folder) {
  files <- sort(list.files(folder, pattern = "_Sholl-Profiles\\.csv$", full.names = TRUE, recursive = TRUE))
  if (!length(files)) return(NULL)
  ids <- vapply(files, extract_sample_name, character(1), folder_name = basename(folder))
  if (any(ids == "") || anyDuplicated(ids)) stop("Duplicate or empty profile IDs in ", folder)
  profiles <- lapply(seq_along(files), function(i) {
    raw <- utils::read.csv(files[i], check.names = FALSE)
    if (!all(c("Radius", "Inters.") %in% names(raw))) stop("Missing Radius or Inters. in ", files[i])
    d <- data.frame(Radius = strict_numeric(raw$Radius, files[i]),
                    Intersections = strict_numeric(raw[["Inters."]], files[i]))
    if (anyNA(d$Radius) || anyDuplicated(d$Radius)) stop("Missing or duplicate radii in ", files[i])
    names(d)[2] <- ids[i]
    d
  })
  combined <- Reduce(function(x, y) merge(x, y, by = "Radius", all = TRUE, sort = TRUE), profiles)
  path <- output_path("sholl_assembly", paste0(basename(folder), "_Sholl_Combined.xlsx"))
  writexl::write_xlsx(list(Sholl = combined, Sources = data.frame(Sample = ids, Source_File = basename(files))), path)
  data.frame(Folder = basename(folder), Profiles = length(files), Output = path)
}
```

```{r run-sholl-assembly, eval=params$run_sholl_assembly}
if (!dir.exists(sholl_parent_dir)) stop("Sholl parent directory not found: ", sholl_parent_dir)
folders <- list.dirs(sholl_parent_dir, recursive = FALSE, full.names = TRUE)
if (!length(folders)) stop("No condition subfolders found.")
assembly_manifest <- dplyr::bind_rows(lapply(folders, combine_sholl_folder))
if (!nrow(assembly_manifest)) stop("No matching Sholl CSV files found.")
print(assembly_manifest)
```

# Reproducibility record

The report records the software environment and enabled analyses. Experimental data
are not bundled with this workflow. An analysis run should retain its configured
notebook, input data, outputs, and session information together.

```{r session-information}
print(params)
print(sessionInfo())
if (any(unlist(params[c("run_viability", "run_bliss", "run_outgrowth", "run_ic50",
                        "run_validation", "run_comparison", "run_sholl_assembly")]))) {
  writeLines(capture.output(sessionInfo()), output_path("", "sessionInfo.txt"))
  settings <- mget(c("params", "viability_files", "treatment_order", "viability_max_time_h",
    "bliss_input_file", "bliss_input_sheet", "bliss_bootstrap_iterations", "bliss_seed",
    "outgrowth_inputs", "outgrowth_doses_uM", "outgrowth_radius_step", "outgrowth_radius_unit",
    "outgrowth_exact", "wilcoxon_adjustment", "plot_error", "single_run_display_cutoff",
    "pooled_display_cutoff", "single_run_min_n", "pooled_min_n", "plot_seed",
    "validation_inputs", "validation_groups", "validation_radius_step", "validation_radius_unit",
    "validation_radius_scale", "comparison_inputs", "comparison_radius_step", "comparison_radius_unit",
    "comparison_radius_scale", "comparison_reference", "comparison_ribbon", "ic50_files",
    "max_IC50", "max_iterations", "sholl_parent_dir"))
  saveRDS(settings, output_path("", "analysis_settings.rds"))
}
```

